\documentclass[11pt]{article}
\usepackage[a4paper,top=2cm,bottom=3cm,left=1.5cm,right=1.5cm]{geometry}
\usepackage{titling}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{thmtools}
\usepackage[shortlabels]{enumitem}
\usepackage{abstract}
\usepackage{hyperref}

\title{Mathematical Analysis of Algorithms}
\date{Spring 2021}

% bold math
\makeatletter
\g@addto@macro\bfseries{\boldmath}
\makeatother

% remove abstract title
\renewcommand{\abstractname}{}
\renewcommand{\absnamepos}{empty}

% style of links
\hypersetup{colorlinks,linkcolor=black}

% set theorem style
\declaretheoremstyle[
  spaceabove=6pt, spacebelow=6pt,
  headfont=\normalfont\bfseries,
  notefont=\normalfont\bfseries,
  bodyfont=\normalfont\upshape,
  postheadspace=0.5em
]{custom}

% set qed symbol
\renewcommand{\qedsymbol}{$\blacksquare$}

% types of theorems
\declaretheorem[style=custom,parent=section]{definition}
\declaretheorem[style=custom,sibling=definition]{example}
\declaretheorem[style=custom,sibling=definition]{theorem}
\declaretheorem[style=custom,sibling=definition]{proposition}
\declaretheorem[style=custom,sibling=definition]{lemma}
\declaretheorem[style=custom,sibling=definition]{corollary}
\declaretheorem[style=custom,sibling=definition]{algorithm}

% use bold fonts to emphasize
\DeclareTextFontCommand{\emph}{\bfseries}

\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}

\begin{document}

% title
\begin{center}
  \LARGE \bfseries \thetitle, \thedate
\end{center}

\begin{abstract}
  This note is taken for the course "Mathematical Analysis of Algorithms", which is instructed by Wen-Chin Chen in Spring 2021.
\end{abstract}

\tableofcontents

\section{February 26, 2021}
\subsection{Course Introduction}
This course focuses on \emph{concrete mathematics}, which is a blend of continuous and discrete mathematics.

\begin{itemize}
  \item Scoring: Homeworks ($100\%$).
  \item Textbook: \textsl{Concrete Mathematics: A Foundation for Computer Science}, by Ronald Graham, Donald Knuth and Oren Patashnik (1989).
\end{itemize}

\subsection{Tower of Hanoi}
The \emph{Tower of Hanoi} consists of three pegs, and we are given $n$ disks, which are initially stacked in decreasing size on one of the three pegs.
The goal is to move the entire tower to one of the other pegs.
In each step, only one disk can be moved, and a larger one cannot be moved onto a smaller one.
We need to determine the minimum number $T_n$ of moves needed to reach the goal.

\begin{theorem}
  $T_n = 2^n - 1$ for all $n \in \NN$.
\end{theorem}
\begin{proof}
  Clearly $T_0 = 0$.
  To move $n$ disks for $n \geq 1$, one ca
  n transfer the $n - 1$ smallest disks to a single peg, then move the largest one to a different peg, and then transfer the $n - 1$ smallest ones back onto the largest one.
  Thus $T_n \leq 2T_{n-1} + 1$ for $n \geq 1$.
  \par Note that we have to move the largest disk at some point, and before that the $n-1$ smallest disks should be transferred to a single peg, requiring $T_{n-1}$ moves.
  Then we move the largest disk.
  While one may move the largest disk more than once, after moving the largest disk for the last time we must transfer the $n-1$ smallest disks back onto the largest one, requiring $T_{n-1}$ moves.
  Thus $T_n \geq 2T_{n-1} + 1$ for $n \geq 1$.
  \par It follows that for $n \geq 1$, we have $T_n = 2T_{n-1} + 1$, implying $T_n + 1 = 2(T_{n-1} + 1)$.
  Thus, $T_n = 2^n(T_0 + 1) - 1 = 2^n - 1$ for all $n \in \NN$.
\end{proof}

\subsection{Josephus Problem}
In the \emph{Josephus problem}, there are $n$ people numbered from $1$ to $n$ standing in a circle.
Every second remaining person is eliminated until only one survives.
We need to determine the number $J_n$ of the survivor.
For instance, we have $J_{10} = 5$ since
\begin{align*}
  (1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
  &\to (3, 4, 5, 6, 7, 8, 9, 10, 1) \\
  &\to (5, 6, 7, 8, 9, 10, 1, 3) \\
  &\to (7, 8, 9, 10, 1, 3, 5) \\
  &\to (9, 10, 1, 3, 5, 7) \\
  &\to (1, 3, 5, 7, 9) \\
  &\to (5, 7, 9, 1) \\
  &\to (9, 1, 5) \\
  &\to (5, 9) \\
  &\to (5).
\end{align*}

\begin{theorem}
  Let $m$ and $\ell$ be integers with $m \geq 0$ and $0 \leq \ell < 2^m$.
  Then $J_{2^m + \ell} = 2\ell + 1$.
\end{theorem}
\begin{proof}
  Clearly $J_1 = 1$.
  We have $J_{2n} = 2J_n - 1$ for $n \geq 1$ since
  \begin{equation*}
    (1, 2, 3, 4, \dots, 2n-1, 2n) \to^n (1, 3, \dots, 2n-1).
  \end{equation*}
  Also we have $J_{2n+1} = 2J_n + 1$ for $n \geq 1$ since
  \begin{equation*}
    (1, 2, 3, 4, \dots, 2n, 2n+1) \to^{n+1} (3, 5, \dots, 2n+1).
  \end{equation*}
  \par Now we show that $J_{2^m + \ell} = 2^\ell + 1$ for $m \geq 0$ and $0 \leq \ell < 2^m$ by induction on $m$.
  For the induction basis, let $m = 0$.
  In this case we must have $\ell = 0$, and $J_{2^0} = 1 = 2 \cdot 0 + 1$ holds.
  Now suppose that $J_{2^{m-1} + \ell} = 2^\ell + 1$ holds for any $0 \leq \ell < 2^{m-1}$.
  If $\ell = 2k$ for some integer $k$, we have
  \begin{equation*}
    J_{2^m + \ell}
    = 2J_{2^{m-1} + k} - 1
    = 2(2k + 1) - 1
    = 2(2k) + 1
    = 2\ell + 1.
  \end{equation*}
  If $\ell = 2k + 1$ for some integer $k$, we have
  \begin{equation*}
    J_{2^m + \ell}
    = 2J_{2^{m-1} + k} + 1
    = 2(2k + 1) + 1
    = 2\ell + 1.
  \end{equation*}
  This completes the proof.
\end{proof}

\section{March 5, 2021}
\subsection{Perturbation Method}
\begin{proposition}
  For $n \geq 1$,
  \begin{equation*}
    \sum_{1 \leq k < n} \sum_{1 \leq j < k} \frac{1}{n-k+j} = n-1.
  \end{equation*}
\end{proposition}
\begin{proof}
  We have
  \begin{align*}
    \sum_{1 \leq k < n} \sum_{1 \leq j < k} \frac{1}{n-k+j}
    &= \sum_{1 \leq k < n} \sum_{1 \leq j < k} \frac{1}{n-j} \\
    &= \sum_{1 \leq j < n} \sum_{j < k \leq n} \frac{1}{n-j} \\
    &= \sum_{1 \leq j < n} 1 \\
    &= n-1.
    \qedhere
  \end{align*}
\end{proof}

\subsection{Finite Calculus}
\begin{theorem}
  Let $(a_k)_{k \geq 0}$ be a sequence.
  Then
  \begin{equation*}
    \sum_{m \leq k < n} (a_{k+1} - a_k) = a_n - a_m
  \end{equation*}
  for $0 \leq m \leq n$.
\end{theorem}
\begin{proof}
  We have
  \begin{align*}
    \sum_{m \leq k < n} (a_{k+1} - a_k)
    &= \sum_{m \leq k < n} a_{k+1} - \sum_{m \leq k < n} a_k \\
    &= \sum_{m+1 \leq k < n+1} a_k - \sum_{m \leq k < n} a_k \\
    &= \Biggl(\sum_{m+1 \leq k < n} a_k + a_n\Biggr) - \Biggl(a_m + \sum_{m+1 \leq k < n} a_k\Biggr) \\
    &= a_n - a_m.
    \qedhere
  \end{align*}
\end{proof}

\begin{lemma}
  Let $n \geq 0$.
  \begin{enumerate}[(a)]
    \item For any real number $c$, $c^{n+1} - c^n = (c-1)c^n$.
    \item For any integer $r \geq 1$,
    \begin{equation*}
      \prod_{0 \leq j < r} (n+1-j) - \prod_{0 \leq j < r} (n-j) = r \prod_{0 \leq j < r-1} (n-j).
    \end{equation*}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \leavevmode
  \begin{enumerate}[(a)]
    \item Trivial.
    \item We have
    \begin{align*}
      \prod_{0 \leq j < r} (n+1-j) - \prod_{0 \leq j < r} (n-j)
      &= (n+1) \prod_{0 \leq j < r-1} (n-j) - (n-r+1) \prod_{0 \leq j < r-1} (n-j) \\
      &= r \prod_{0 \leq j < r-1} (n-j).
      \qedhere
    \end{align*}
  \end{enumerate}
\end{proof}

\begin{theorem}
  Let $n \geq 0$.
  \begin{enumerate}[(a)]
    \item For any real number $c \neq 1$,
    \begin{equation*}
      \sum_{m \leq k < n} c^k = \frac{c^n - c^m}{c - 1}.
    \end{equation*}
    \item For any integer $r \geq 0$,
    \begin{equation*}
      \sum_{m \leq k < n} \prod_{0 \leq j < r} (k-j) = \frac{1}{r+1} \Biggl(\prod_{0 \leq j < r+1} (n-j) - \prod_{0 \leq j < r+1} (m-j)\Biggr).
    \end{equation*}
  \end{enumerate}
\end{theorem}
\begin{proof}
  \leavevmode
  \begin{enumerate}[(a)]
    \item We have
    \begin{equation*}
      \sum_{m \leq k < n} c^k
      = \frac{1}{c-1} \sum_{m \leq k < n} (c^{k+1} - c^k)
      = \frac{c^n - c^m}{c-1}.
    \end{equation*}
    \item We have
    \begin{align*}
      \sum_{m \leq k < n} \prod_{0 \leq j < r} (k-j)
      &= \frac{1}{r+1} \sum_{m \leq k < n} \Biggl(\prod_{0 \leq j < r+1} (k+1-j) - \prod_{0 \leq j < r+1} (k-j)\Biggr) \\
      &= \frac{1}{r+1} \Biggl(\prod_{0 \leq j < r+1} (n-j) - \prod_{0 \leq j < r+1} (m-j)\Biggr).
      \qedhere
    \end{align*}
  \end{enumerate}
\end{proof}

\begin{theorem}[Summation by Parts]
  Let $(a_n)_{n \geq 0}$ and $(b_n)_{n \geq 0}$ be sequences.
  Then for $0 \leq m \leq n$,
  \begin{equation*}
    \sum_{m \leq k < n} a_k(b_{k+1} - b_k) = (a_nb_n - a_mb_m) - \sum_{m \leq k < n} (a_{k+1}- a_k)b_{k+1}.
  \end{equation*}
\end{theorem}
\begin{proof}
  We have
  \begin{align*}
    \sum_{m \leq k < n} a_k(b_{k+1} - b_k)
    &= \sum_{m \leq k < n} \Bigl((a_{k+1}b_{k+1} - a_kb_k) - (a_{k+1} - a_k)b_{k+1}\Bigr) \\
    &= (a_nb_n - a_mb_m) - \sum_{m \leq k < n} (a_{k+1} - a_k)b_{k+1}.
    \qedhere
  \end{align*}
\end{proof}

\begin{corollary}
  For $n \geq 1$,
  \begin{equation*}
    \sum_{1 \leq k < n} H_k = n(H_n - 1).
  \end{equation*}
\end{corollary}
\begin{proof}
  We have
  \begin{align*}
    \sum_{1 \leq k < n} H_k
    &= (nH_n - 1) - \sum_{1 \leq k < n} (H_{k+1} - H_k)(k+1) \\
    &= (nH_n - 1) - (n - 1) \\
    &= n(H_n - 1).
    \qedhere
  \end{align*}
\end{proof}

\section{March 12, 2021}
\subsection{Floor and Ceiling Sums}
\begin{proposition}
  Let $m$ be a positive integer.
  Then for any real number $x$, we have
  \begin{equation*}
    \sum_{0 \leq k < m} \biggl\lfloor \frac{x + k}{m} \biggr\rfloor = \lfloor x \rfloor.
  \end{equation*}
\end{proposition}
\begin{proof}
  Let $r = m - \lfloor x \rfloor \bmod m$.
  Then we have
  \begin{align*}
    \sum_{0 \leq k < m} \biggl\lfloor \frac{x + k}{m} \biggr\rfloor
    &= \sum_{0 \leq k < m} \biggl\lfloor \frac{m \lfloor x/m \rfloor + x \bmod m + k}{m} \biggr\rfloor \\
    &= \sum_{0 \leq k < m} \biggl( \biggl\lfloor \frac{x}{m} \biggr\rfloor + \biggl\lfloor \frac{x \bmod m + k}{m} \biggr\rfloor \biggr) \\
    &= m \biggl\lfloor \frac{x}{m} \biggr\rfloor + m - r \\
    &= m \biggl\lfloor \frac{x}{m} \biggr\rfloor + \lfloor x \rfloor \bmod m \\
    &= \lfloor x \rfloor.
    \qedhere
  \end{align*}
\end{proof}

\begin{theorem}
  Let $m$ be a positive integer and let $n$ be an integer.
  Let $d = \gcd(n, m)$.
  Then for any real number $x$, we have
  \begin{equation*}
    \sum_{0 \leq k < m} \biggl\lfloor \frac{x + kn}{m} \biggr\rfloor = d\biggl\lfloor\frac{x}{d}\biggr\rfloor + \frac{(m-1)(n-1)}{2} + \frac{d-1}{2}.
  \end{equation*}
\end{theorem}
\begin{proof}
  We have
  \begin{align*}
    \sum_{0 \leq k < m} \biggl\lfloor \frac{x + kn}{m} \biggr\rfloor
    &= \sum_{0 \leq k < m} \biggl\lfloor \frac{x + kn \bmod m}{m} \biggr\rfloor + \sum_{0 \leq k < m} \frac{kn}{m} - \sum_{0 \leq k < m} \frac{kn \bmod m}{m} \\
    &= d \sum_{0 \leq k < m/d} \biggl\lfloor \frac{x + kd}{m} \biggr\rfloor + \sum_{0 \leq k < m} \frac{kn}{m} - d \sum_{0 \leq k < m/d} \frac{kd}{m} \\
    &= d \biggl\lfloor \frac{x}{d} \biggr\rfloor + \frac{n(m-1)}{2} - \frac{m - d}{2} \\
    &= d \biggl\lfloor \frac{x}{d} \biggr\rfloor + \frac{(m-1)(n-1)}{2} + \frac{d-1}{2}.
    \qedhere
  \end{align*}
\end{proof}

\section{March 16, 2021}
\subsection{Congruence}
\begin{definition}
  Let $m$ be a positive integer.
  For $a, b \in \ZZ$, we write $a \equiv b \pmod m$ if $m \mid a - b$.
\end{definition}

\begin{proposition}
  Let $m$ be a positive integer.
  Then the following properties hold for any $a, b, c \in \ZZ$.
  \begin{enumerate}[(a)]
    \item $a \equiv a \pmod m$.
    \item If $a \equiv b \pmod m$, then $b \equiv a \pmod m$.
    \item If $a \equiv b \pmod m$ and $b \equiv c \pmod m$, then $a \equiv c \pmod m$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  \leavevmode
  \begin{enumerate}[(a)]
    \item We have $a \equiv a \pmod m$ since $m \mid 0 = a - a$.
    \item Suppose that $a \equiv b \pmod m$, i.e., $m \mid a - b$.
    Then $m \mid -(a - b) = b - a$.
    Thus, $b \equiv a \pmod m$.
    \item Suppose that $a \equiv b \pmod m$ and $b \equiv c \pmod m$, i.e., $m \mid a - b$ and $m \mid b - c$.
    Then
    \begin{equation*}
      m \mid (a - b) + (b - c) = a - c.
    \end{equation*}
    Thus, $a \equiv c \pmod m$.
    \qedhere
  \end{enumerate}
\end{proof}

\begin{proposition}
  Let $m$ be a positive integer.
  Then the following properties hold for any $a, b, c, d \in \ZZ$.
  \begin{enumerate}[(a)]
    \item If $a \equiv b \pmod m$ and $c \equiv d \pmod m$, then $a + c \equiv b + d \pmod m$.
    \item If $a \equiv b \pmod m$ and $k \in \ZZ$, then $ka \equiv kb \pmod m$.
    \item If $a \equiv b \pmod m$ and $c \equiv d \pmod m$, then $ac \equiv bd \pmod m$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  \leavevmode
  \begin{enumerate}[(a)]
    \item Since $m \mid a - b$ and $m \mid c - d$, we have $m \mid (a - b) + (c - d) = (a + c) - (b + d)$.
    \item Since $m \mid a - b$, we have $m \mid k(a - b) = ka - kb$.
    \item We have $ac \equiv bc \equiv bd \pmod m$.
    \qedhere
  \end{enumerate}
\end{proof}

\begin{definition}
  Let $m$ be a positive integer.
  For any integer $a$, let $[a]_m = \{b \in \ZZ: a \equiv b \pmod m\}$.
\end{definition}

\section{March 19, 2021}

\section{March 26, 2021}
\subsection{Stern--Brocot Tree}
The \emph{Stern--Brocot tree} is an infinite binary tree satisfying the following properties.
\begin{enumerate}[(i)]
  \item The root of Stern--Brocot tree contains the identity matrix
  \begin{equation*}
    I = \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}.
  \end{equation*}
  \item For each node containing $A$, its left child and its right child contain $AL$ and $AR$, respectively, where
  \begin{equation*}
    L = \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}
    \qquad \text{and} \qquad
    R = \begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}.
  \end{equation*}
\end{enumerate}
For each node containing matrix
\begin{equation*}
  A = \begin{pmatrix} n & n' \\ m & m' \end{pmatrix},
\end{equation*}
we also define its corresponding fraction as
\begin{equation*}
  q = \frac{m + m'}{n + n'}.
\end{equation*}

\subsection{Permutations and Combinations}
\begin{itemize}
  \item The number of permutations to choose $k$ elements from a collection of $n$ distinct items equals to
  \begin{equation*}
    P(n, k) = n^{\underline{k}} = n(n-1) \cdots (n-k+1).
  \end{equation*}
  \item The number of permutations to choose $k$ elements from $n$ types of items equals to
  \begin{equation*}
    U(n, k) = n^k.
  \end{equation*}
  \item The number of combinations to choose $k$ elements from a collection of $n$ distinct items equals to
  \begin{equation*}
    C(n, k) = \binom{n}{k} = \frac{n(n-1) \cdots (n-k+1)}{k(k-1) \cdots 1}.
  \end{equation*}
  \item The number of combinations to choose $k$ elements from $n$ types of items equals to
  \begin{equation*}
    H(n, k) = \biggl(\!\!\binom{n}{k}\!\!\biggr) = \frac{(n+k-1)(n+k) \cdots n}{k(k-1) \cdots 1}.
  \end{equation*}
\end{itemize}

We have the following recurrence relations.
\begin{align*}
  P(n, k) &= P(n-1, k) + kP(n-1, k-1), \\
  U(n, k) &= nU(n, k-1), \\
  C(n, k) &= C(n-1, k) + C(n-1, k-1), \\
  H(n, k) &= H(n-1, k) + H(n, k-1).
\end{align*}

\section{March 30, 2021}
\subsection{Binomial Coefficients}
\begin{proposition}
  We have the following identities.
  \begin{enumerate}[(a)]
    \item
    \begin{equation*}
      \binom{n}{k} = \binom{n}{n-k}.
    \end{equation*}
    \item
    \begin{equation*}
      k\binom{r}{k} = r\binom{r-1}{k-1}.
    \end{equation*}
    \item
    \begin{equation*}
      \binom{r}{k} = \binom{r-1}{k} + \binom{r-1}{k-1}.
    \end{equation*}
    \item
    \begin{equation*}
      \binom{r}{k} = (-1)^k\binom{k-1-r}{k}.
    \end{equation*}
    \item
    \begin{equation*}
      \sum_{k \geq 0} \binom{r}{k} x^k y^{r-k} = (x+y)^r.
    \end{equation*}
    \item
    \begin{equation*}
      \sum_{0 \leq k \leq n} \binom{k}{m} = \binom{n+1}{m+1}.
    \end{equation*}
    \item
    \begin{equation*}
      \sum_{0 \leq k \leq n} \binom{r+k}{k} = \binom{n+1+r}{n}.
    \end{equation*}
    \item
    \begin{equation*}
      \sum_{0 \leq k \leq n} \binom{r}{k}\binom{s}{n-k} = \binom{r+s}{n}.
    \end{equation*}
  \end{enumerate}
\end{proposition}

\begin{example}
  Let us try to compute
  \begin{equation*}
    \sum_{0 \leq k \leq n} k\binom{m-k-1}{m-n-1} \Bigm/ \binom{m}{n}
  \end{equation*}
  for $n, m \in \ZZ$.
  First, notice that
  \begin{equation*}
    \sum_{0 \leq k \leq n} k\binom{m-k-1}{m-n-1}
    = \sum_{0 \leq k \leq n} m\binom{m-k-1}{m-n-1} - \sum_{0 \leq k \leq n} (m-k)\binom{m-k-1}{m-n-1}.
  \end{equation*}
  We have
  \begin{equation*}
    \sum_{0 \leq k \leq n} \binom{m-k-1}{m-n-1} = \binom{m}{m-n} = \binom{m}{n}
  \end{equation*}
  and
  \begin{equation*}
    \sum_{0 \leq k \leq n} \frac{m-k}{m-n}\binom{m-k-1}{m-n-1} = \sum_{0 \leq k \leq n} \binom{m-k}{m-n} = \binom{m+1}{m-n+1} = \frac{m+1}{m-n+1}\binom{m}{n}.
  \end{equation*}
  Thus,
  \begin{equation*}
    \sum_{0 \leq k \leq n} k\binom{m-k-1}{m-n-1} \Bigm/ \binom{m}{n}
    = m - \frac{(m-n)(m+1)}{m-n+1}
    = \frac{n}{m-n+1}.
  \end{equation*}
\end{example}

\begin{example}
  Let us try to compute
  \begin{equation*}
    \sum_{k \geq 0} \binom{n+k}{2k} \binom{2k}{k} \frac{(-1)^k}{k+1}
  \end{equation*}
  for $n \in \NN$.
\end{example}

\section{April 9, 2021}
\subsection{Hypergeometric Functions}
The \emph{generalized hypergeometric series} is a series
\begin{equation*}
  F(a_1, a_2, \dots, a_m; b_1, b_2, \dots, b_n; z) = \sum_{k \geq 0} t_k
\end{equation*}
such that $t_0 = 1$ and
\begin{equation*}
  \frac{t_{k+1}}{t_k} = \frac{(k + a_1)(k + a_2) \cdots (k + a_m)z}{(k + b_1)(k + b_2) \cdots (k + b_n)(k + 1)}
\end{equation*}
for each integer $k \geq 0$, i.e., we have
\begin{equation*}
  F(a_1, a_2, \dots, a_m; b_1, b_2, \dots, b_n; z) = \sum_{k \geq 0} \frac{a_1^{\overline{k}}a_2^{\overline{k}} \cdots a_m^{\overline{k}}}{b_1^{\overline{k}}b_2^{\overline{k}} \cdots b_n^{\overline{k}}} \frac{z^k}{k!}.
\end{equation*}

\end{document}
